### 2020-02-13 report.tex
some edits

There are at least two possible side-effects of chronological backtracking which may explain its effectiveness. It has been observed that when a solver backtracks non-chronologically after a conflict, many of the literals that are unassigned during the backtrack will be re-assigned again in roughly the same order when the solver continues. By using chronological backtracking, the solver keeps a lot more of its partial assignment intact and saves from having to reconstruct a lot of the trail via propagation that it otherwise would have. Saving from having to reconstruct portions of the trail after backtracking was first introduced in the context of restarts \cite{DBLP:journals/jsat/TakRH11}. The second side-effect of chronological backtracking is that it forces the solver to stay in the same local neighbourhood during search, since it unassigns less of the partial assignment and one could argue that this might lead to finding additional conflicts more quickly or reaching a satisfying assignment (if one exists) more quickly. This second side-effect also leads to the search with chronological backtracking being less flexibile to jump out of local neighbourhoods and obey the variable selection heuristic. We argue that this second side-effect is not the primary explanation for the effectiveness of chronological backtracking, and aim to to develop a technique which will produce a benefit similar to the first side-effect while maintaining the freedom for the solver to jump out of local neighbourhoods and obey the variable selection heuristic.

@@

@@

\begin{figure}\includegraphics[scale=0.8]{trail_diagram.pdf}\caption{Storing into a trail cache after backtracking.}\end{figure}

@@

When concatenating trail caches together as above, it is important to note that the running trail cache can grow indefinitely. The trail cache can be trimmed periodically in the following way. Traverse down the trail cache and whenever a literal appears more than once, delete its entry in the trail cache. Whenever a literal $x$ appears whose complement $\lnot x$ has already appeared on the trail, $x$ should be deleted and everything remaining until the end of the trail cache should be deleted (in order to keep the trail cache free of conflicts).  We decided to trim the trail cache whenever its size exceeded twice the total number of variables, as that is what worked best in preliminary experiments.\newline

@@

\begin{figure}
\begin{tabular}{|l|l|l|l|l|l|}
\hline
                         & Total Solved & SAT & UNSAT \\ \hline
cadical                  & 492          & 289 &  203  \\ \hline
cadical\_chrono          & 498          & 292 &  206  \\ \hline
cadical\_trail           & 493          & 290 &  203  \\ \hline
cadical\_trail\_enhanced & 496          & 292 &  204  \\ \hline
\end{tabular}
\caption{Table of results for cadical, version pulled from github as of July 30, 2020.}
\end{figure}

@@

With more clever data structures the efficiency of restoring the trail could be improved. There are a number of parameters which can be tuned, including how often to refresh the trail, how large the cutoff point for rejecting "long" reasons should be, how long to wait to prune the trail cache if one is appending multiple trail caches together, how far to look ahead on the trail cache for conflicts (in terms of number of decisions or total number of literals). We only experimented with a few different settings of these parameters that made sense, but more careful fine-tuning of these parameters could further enhance the effectiveness of the technique. There are potentially other techniques one could use a trail cache for, including inprocessing steps that involve probing. A look down the trail cache is an incomplete but very fast way to do a probing step, as opposed to propagation. Trail saving and chronological backtracking could be combined in some way to produce even better results.
\bibliography{ref}{}
\bibliographystyle{splncs04}
\end{document}

>>>>

There are at least two possible side-effects of chronological backtracking which may explain its effectiveness. It has been observed that when a solver backtracks non-chronologically after a conflict, many of the literals that are unassigned during the backtrack will be re-assigned again in roughly the same order when the solver continues. By using chronological backtracking, the solver keeps a lot more of its partial assignment intact and saves from having to reconstruct a lot of the trail via propagation that it otherwise would have. Saving from having to reconstruct portions of the trail after backtracking was first introduced in the context of restarts \cite{DBLP:journals/jsat/TakRH11}. The second side-effect of chronological backtracking is that it forces the solver to stay in the same local neighbourhood during search, since it unassigns less of the partial assignment and one could argue that this might lead to finding additional conflicts more quickly or reaching a satisfying assignment (if one exists) more quickly. However, this second side-effect also leads to less search flexibility, since the solver does not jump out of local neighbourhoods as often and does not obey the variable selection heuristic. We will argue that it is the first side-effect that leads to the effectiveness of chronological backtracking and not the second by developing a technique which also saves from having to repeat propagations, but allows the search to remain flexible and obey the variable selection heuristic.

@@

\begin{figure}\center\includegraphics[scale=0.7]{trail_diagram.pdf}\caption{Storing into a trail cache after backtracking.}\end{figure}

@@

@@

When concatenating trail caches together as above, it is important to note that the running trail cache can grow indefinitely. The trail cache can be trimmed periodically in the following way. Traverse down the trail cache and whenever a literal appears more than once, delete its entry in the trail cache. Whenever a literal $x$ appears whose complement $\lnot x$ has already appeared on the trail cache, $x$ should be deleted and everything remaining until the end of the trail cache should be deleted (in order to keep the trail cache free of conflicts).  We decided to trim the trail cache whenever its size exceeded twice the total number of variables in the SAT instance, as that is what worked best in preliminary experiments.\newline

@@

\begin{figure}
\begin{tabular}{|l|l|l|l|l|l|}
\hline
                         & Total Solved & SAT & UNSAT & Par-2 (x 10\textasciicircum{}6 s) \\ \hline
cadical                  & 492          & 289 &  203  & 3.5020                            \\ \hline
cadical\_chrono          & 498          & 292 &  206  & 3.4400                            \\ \hline
cadical\_trail           & 493          & 290 &  203  & 3.5228                            \\ \hline
cadical\_trail\_enhanced & 496          & 292 &  204  & 3.4847                            \\ \hline
\end{tabular}
\caption{Table of results for cadical, version pulled from github as of July 30, 2020.}
\end{figure}

@@

With more clever data structures the efficiency of restoring the trail could be improved. There are a number of parameters which can be tuned, including how often to refresh the trail, how large the cutoff point for rejecting "long" reasons should be, how long to wait to prune the trail cache if one is appending multiple trail caches together, how far to look ahead on the trail cache for conflicts (in terms of number of decisions or total number of literals). We only experimented with a few different settings of these parameters that made sense, but more careful fine-tuning of these parameters could further enhance the effectiveness of the technique. There are potentially other techniques one could use a trail cache for, including inprocessing steps that involve probing. A look down the trail cache is an incomplete but very fast way to do a probing step, as opposed to propagation. It is also possible that trail saving and chronological backtracking could be combined in some way to produce even better results.
\bibliography{ref}{}
\bibliographystyle{splncs04}
\end{document}
